\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-31 analysis/paper/paper.Rmd
fixed penalized issue.

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

@@

\textbf{casebase}        & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{CFC}             & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\ \bottomrule
\textbf{cmprsk}             & x                        &                           &                       &                  &                     &  x                        &                                  & x                                \\ \bottomrule
\textbf{CoxRidge}        &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{crrp}            & x                        &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{fastcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{flexrsurv}       &                          & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{flexsurv}        & x                        & x                         &                       & x                & x                   &                          &                                  & x               \\ \bottomrule
\textbf{glmnet}          &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{glmpath}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{mets}            & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{penalized}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\ \bottomrule
\textbf{riskRegression}  & x                         &                           & x                     &                  &                     & x                        &                                  & x                                \\ \bottomrule
\textbf{rstpm2}          &                          & x                         &                      & x                & x                   & x                        & x                                & x                         \\ \bottomrule
\textbf{SmoothHazard}    &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\ \bottomrule
\textbf{survival}        & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\ \bottomrule

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features. Competing Risks: whether or not an implementation for competing risks is present. Allows Non-PH: permits models for non-proportional hazards. Penalized Regression: allows for a penalty term when estimating hazards (e.g. lasso or ridge). Splines: permits a flexible fit on time through the use of splines. Parametric: implementation for parametric models. Semi-parametric: implementation for semi-parametric models. Interval/left censoring: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. Absolute Risk: computation for survival curves, cumulative incidence or cumulative hazard is available.}

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the Weibull model is just as good as the larger (in terms of number of parameters) spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fits$weibull, fits$splines, test = "LRT")
```

@@

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95\\% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

@@

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model. As we can see, the two approaches agree quite well for AML; however, there seems to be a difference of about 5% between the two curves for ALL. This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).
# Case study 3---SUPPORT Data

@@

In the original study, the authors developed two scores for predicting the outcome. These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}). Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}. We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used to assess both discrimination and calibration in a single model. Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models. To account for overfitting, we will use 95% of the observations to train our models, while the remaining 5% will be used to compute absolute risk curves and the Brier scores.

```{r supportData, eval = eval_cs3}
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

In Figure\ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves. On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall. For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.

<!--Before generating an absolute risk curve using regularized logistic regression, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized logistic regression and the unadjusted Kaplan-Meier curve.-->

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme
```

@@

Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data. We first create a matrix \code{newx} containing the relevant variables. Using \code{times}, we specify the time points at which the cumulative incidence will be computed. Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.

```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
# Estimating the absolute risk curve using the newdata parameter
newx <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = test)[, -c(1)]
times <- sort(unique(test$d.time))

@@

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized logistic"

ab <- rbind(abCB, abCN)
```

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized logistic model, aside from cancer state (cano).

@@

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M","Cox","Penalized Cox","Penalized logistic")) + 
  paper_gg_theme
```

@@

  # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
  # the output of absoluteRisk is an array with dimension dependening on the length of the requested times:
  # case 1: the number of time points is 1
  #         dim(array) =  (length(time), NROW(newdata), number of causes in the data)
  if (length(times) == 1) {
    a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
    p <- matrix(a, ncol = 1)
  } else {
    # case 2 a) zero is included in the number of time points
    if (0 %in% times) {
      # dim(array) =  (length(time)+1, NROW(newdata)+1, number of causes in the data)
      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
      p <- t(a)
    } else {
      # case 2 b) zero is not included in the number of time points (but the absoluteRisk function adds it)
      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
      ### we need to invert the plot because, by default, we get cumulative incidence
      #a[, -c(1)] <- 1 - a[, -c(1)]
      ### we remove time 0 for everyone, and remove the time column
      a <- a[-c(1), -c(1)] ### a[-c(1), ] to keep times column, but remove time 0 probabilities
      # now we transpose the matrix because in riskRegression we work with number of
      # observations in rows and time points in columns
      p <- t(a)
    }
  }
  if (NROW(p) != NROW(newdata) || NCOL(p) != length(times)) {
    stop(paste("\nPrediction matrix has wrong dimensions:\nRequested newdata x times: ", 
               NROW(newdata), " x ", length(times), "\nProvided prediction matrix: ", 
               NROW(p), " x ", NCOL(p), "\n\n", sep = ""))
  }
  p
}
```

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. All follow-up times seen in the new data were used to generate the curves. The Cox-based curves appear to be smooth, but this is due to the 1703 follow-up times seen while fitting the models and the 319 time points used to estimate the curves. (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. Lower is better. Penalized logistic regression produced the lowest Brier scores over follow-up time.", warning = FALSE}
# Compute Brier score for penalized models
brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized Logistic" = cbFull),
                           data = cbind(subset(test, select = c(d.time, death)),
                                        as.data.frame(newx)), 
                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
#brierFinalResults$models<-as.character(brierFinalResults$models)
# Combine Brier scores between unpenalized and penalized models before plotting

data_brier <- bind_rows(
  brierFinalResults$Brier$score,
  filter(brierFull$Brier$score, model == "PredictionModel")
  )
data_brier$model<-factor(data_brier$model,c("Null model", "PredictionModel","Penalized Cox", "Penalized Logistic"))
levels(data_brier$model)<-c("Kaplan-Meier","Cox","Penalized Cox","Penalized logistic")

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (Years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4)

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

\toprule

@@

\bottomrule
\textbf{casebase}        & \checkmark                        & \checkmark                         & \checkmark                     & \checkmark                & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{CFC}             & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   &                          &                                  & \checkmark                                \\ \hline
\textbf{cmprsk}             & \checkmark                        &                           &                       &                  &                     &  \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{CoxRidge}        &                          & \checkmark                         & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{crrp}            & \checkmark                        &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{fastcox}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{flexrsurv}       &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{flexsurv}        & \checkmark                        & \checkmark                         &                       & \checkmark                & \checkmark                   &                          &                                  & \checkmark               \\ \hline
\textbf{glmnet}          &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{glmpath}         &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{mets}            & \checkmark                        &                           &                       & \checkmark                &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{penalized}       &                          &                           & \checkmark                     &                  &                     & \checkmark                        &                                  &                                  \\ \hline
\textbf{riskRegression}  & \checkmark                         &                           & \checkmark                     &                  &                     & \checkmark                        &                                  & \checkmark                                \\ \hline
\textbf{rstpm2}          &                          & \checkmark                         &                      & \checkmark                & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                         \\ \hline
\textbf{SmoothHazard}    &                          & \checkmark                         &                       & \checkmark                & \checkmark                   &                          & \checkmark                            &                                      \\ \hline
\textbf{survival}        & \checkmark                        & \checkmark                         &                       &                  & \checkmark                   & \checkmark                        & \checkmark                                & \checkmark                               \\ 
\bottomrule

@@

\caption{Comparison of various \proglang{R} packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Absolute Risk}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from \pkg{casebase} on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on our earlier discussion of the population-time plot, this poor fit for the exponential hazard was expected. We notice a better fit with increasing complexity of our model for time in Figure \ref{fig:erspc-cox-cif} (panels 2--4). As noted above, the usual asymptotic results hold for likelihood ratio tests built using case-base sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) spline model:

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fits$exponential, fits$splines, test = "LRT")
```

@@

In Figure \ref{fig:interaction-ERSPC} we plot the estimated death from prostate cancer hazard ratio and 95% confidence interval for screening vs. control group as a function of time using the \code{plot} method for objects of class \code{singleEventCB}:

@@

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both case-base sampling and the Fine-Gray model. As we can see, the two approaches agree quite well for AML; however, there seems to be a difference of about 5% between the two curves for ALL. This difference does not appear to be significant: the curve from case-base sampling is contained within a 95% confidence band around the Fine-Gray absolute risk curve (figure not shown).

# Case study 3---SUPPORT Data

@@

In the original study, the authors developed two scores for predicting the outcome. These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}). Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}. We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used to assess both discrimination and calibration in a single model. Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models. We will use 95% of the observations to train our models, and the remaining 5% to compute absolute risk curves and Brier scores.

```{r supportData, eval = eval_cs3}
scaleFUNy <- function(x) sprintf("%.2f", x)
scaleFUNx <- function(x) sprintf("%.0f", x)
data(support)
# Change time to years
support$d.time <- support$d.time/365.25

@@

In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves. On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall. For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.

<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->

@@

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme + 
  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
```

@@

Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data. We first create a matrix \code{newx} containing the relevant variables. Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed. Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.

```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
# Estimating the absolute risk curve using the newdata parameter
newx <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = test)[, -c(1)]
times <- sort(unique(test$d.time))
#times<-seq(from=min(times),to=max(times),length.out = 100)

@@

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized CB"

ab <- rbind(abCB, abCN)

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).

@@

These values do not inform us of model performance. To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time. The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). There is separation between the curves, but it is minimal. To better demonstrate the differences in performance between the models, we compare Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models. The differences, however, are minimal. All three outperform the Kaplan-Meier estimator.

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M", "Cox", "Penalized Cox", "Penalized CB")) + 
  paper_gg_theme + 
  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 9)
```

@@

  # if (missing(cause)) stop("Argument cause should be the event type for which we predict the absolute risk.")
  # the output of absoluteRisk is an array with dimension depending on the length of the requested times:
  # case 1: the number of time points is 1
  #         dim(array) =  (length(time), NROW(newdata), number of causes in the data)
  if (length(times) == 1) {
    a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
    p <- matrix(a, ncol = 1)
  } else {
    # case 2 a) zero is included in the number of time points
    if (0 %in% times) {
      # dim(array) =  (length(time)+1, NROW(newdata)+1, number of causes in the data)
      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
      p <- t(a)
    } else {
      # case 2 b) zero is not included in the number of time points (but the absoluteRisk function adds it)
      a <- casebase::absoluteRisk(object, newdata = newdata, time = times)
      ### we need to invert the plot because, by default, we get cumulative incidence
      #a[, -c(1)] <- 1 - a[, -c(1)]
      ### we remove time 0 for everyone, and remove the time column
      a <- a[-c(1), -c(1)] ### a[-c(1), ] to keep times column, but remove time 0 probabilities
      # now we transpose the matrix because in riskRegression we work with number of
      # observations in rows and time points in columns
      p <- t(a)
    }
  }
  if (NROW(p) != NROW(newdata) || NCOL(p) != length(times)) {
    stop(paste("\nPrediction matrix has wrong dimensions:\nRequested newdata x times: ", 
               NROW(newdata), " x ", length(times), "\nProvided prediction matrix: ", 
               NROW(p), " x ", NCOL(p), "\n\n", sep = ""))
  }
  p
}
```

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
# Compute Brier score for penalized models
brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
                           data = cbind(subset(test, select = c(d.time, death)),
                                        as.data.frame(newx)), 
                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
#brierFinalResults$models<-as.character(brierFinalResults$models)
# Combine Brier scores between unpenalized and penalized models before plotting

data_brier <- bind_rows(
  brierFinalResults$Brier$score,
  filter(brierFull$Brier$score, model == "PredictionModel")
  ) %>% 
  mutate(model = as.character(model), # First return to string, so types match
         model = case_when(
    model == "Null model" ~ "K-M",
    model == "PredictionModel" ~ "Cox",
    TRUE ~ model
  ),
  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized CB", "K-M"),
                 labels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4) + 
  scale_x_continuous(labels = scaleFUNx, 
                     breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
